---
title: "Breast cancer survival analysis with R"
#author: "Melvin Galera"
#date: "2024-03-26"
output:
  github_document:
    toc: false
---

```{r setup, include=FALSE}
# Global code chunk options; adjust individual codes as required
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.width = 12, fig.height = 12)

# Load libraries
library(tidyverse)
library(survival)
library(ggpubr)
library(survminer)
library(sjPlot)
library(finalfit)


# Load data
cancer_df <-survival::rotterdam 

```

<br>

### I. Project objective

Breast cancer is the most common cancer diagnosed in women and is the second most common cause of cancer death among women worldwide. The primary treatment recommended for most types of breast cancer is surgery because it is the most effective means of breast cancer removal when it is technically feasible to remove it, and recommended in all stages except the last. 

This project was aimed at performing a survival analysis for primary breast cancer patients who have undergone breast cancer surgery based on several factors or variables (patient's characteristics, tumor characteristics, and treatment received) and provide estimates of each of the variable's hazard ratios.  

<br>

### II. Exploratory Data Analysis

**A. Data Collection**

In this project, the in-built dataset "rotterdam" from the survival package in R was used. This dataset has 2982 records of patients with primary breast cancers available in the Rotterdam tumor bank. Primary breast cancer which is breast cancer that has not spread beyond the breast or the lymph nodes under the arm. The dataset has 15 variables/covariates:

  * `pid`     : patient identifier
  * `year`    : year of surgery   
  * `age`     : patient's age at surgery
  * `meno`    : patient's menopausal status (0 = premenopausal, 1 = postmenopausal)
  * `size`    : the range of tumor sizes (<=20, 20-50,  50<)
  * `grade`   : differentiation grade of tumor
  * `nodes`   : number of lymph nodes noted to be positive
  * `pgr`     : biomarker of progesterone receptor (fmol/l)
  * `er`      : biomarker of estrogen receptor (fmol/l)
  * `hormon`  : indicates whether hormonal treatment was received or not (0 = no, 1 = yes)
  * `chemo`   : indicates whether chemotherapy was received or not 
  * `rtime``  : days to relapse or lost-to-follow-up
  * `recur``  : indicates whether the patient has had a recurrence or not (0 = no relapse, 1 = relapse)
  * `dtime`   : duration of survival or days to death or last follow up
  * `death`   : indicates whether the patient is alive or not (0 = alive, 1 = dead)
  
<br>

**B. Data Exploration**

An initial look at the structure of the dataset (`cancer_df`) shows:

```{r}
str(cancer_df)
```
The first 10 records of `cancer_df`:

```{r}
cancer_df %>% head(10)
```

```{r, include=FALSE}
# check for missing data
colSums(is.na(cancer_df))
```

```{r}
age.breaks <- c(24, 61, 91)
age.tags <- c("24-60", "61-90")
cancer_df$age.group <- cut(cancer_df$age, breaks = age.breaks,
                           include.lowest = TRUE,
                           right = FALSE,
                           labels = age.tags)

nodes.breaks <- c(0, 11, 21, 35)
nodes.tags <- c("0-10", "11-20", "21-34")
cancer_df$nodes.group <- cut(cancer_df$nodes, breaks = nodes.breaks,
                           include.lowest = TRUE,
                           right = FALSE,
                           labels = nodes.tags)
```


```{r}
cancer_df <- cancer_df %>% 
  mutate(
    gradeF = factor(grade),
    menoF = factor(meno, labels = c("No", "Yes")), 
    hormonF = factor(hormon, labels = c("No", "Yes")), 
    chemoF = factor(chemo, labels = c("No", "Yes")),
    recurF = factor(recur, labels = c("No", "Yes")),
    deathF = factor(death, labels = c("No", "Yes"))
  )
```

<br>

Values of some features were grouped as factors. The feature `age` was factored as `age.group` with levels "24-60" and "61-90".  The feature `nodes` was factored as `nodes.group` with levels "0-10", "11-20", and "21-34". Other features such as `grade`, `meno`, `hormon`, `chemo`, `recur`, and `death` were also factored into levels "no" and "yes" (see the 'F' added in the names to indicate factored).

Summary statistics of the resulting relevant features are shown below. There are no missing data in each feature.

```{r}
cancer_df_stat <- cancer_df %>% 
  select(-c(pid, grade, meno, hormon, chemo, recur, death))

summary(cancer_df_stat %>% select(year, age, age.group, nodes, nodes.group))
```
```{r}
summary(cancer_df_stat %>% select(pgr, er, size, gradeF, menoF))
```
```{r}
summary(cancer_df_stat %>% select(hormonF, chemoF, recurF, deathF))
```

<br>

**Univariate plots**

```{r, fig.align='center', fig.width=15, fig.height=5}

p1_age <- ggplot(data = cancer_df, aes(x = age)) +
  geom_histogram(binwidth= 7, color= "#ce90b4", fill = "#b4006a") +
  theme(plot.title = element_text(size = 16), axis.title = element_text(size = 16), axis.text = element_text(size = 16)) +
  labs(x = "Age",
       y = "Count",
       title = "Age distribution")

p2_nodes <- ggplot(data = cancer_df, aes(x = nodes)) +
  geom_histogram(binwidth= 3, color= "#ce90b4", fill = "#b4006a") +
  theme(plot.title = element_text(size = 16), axis.title = element_text(size = 16), axis.text = element_text(size = 16)) +
  labs(x = "Nodes",
       y = "Count",
       title = "Nodes distribution")

p3_pgr <- ggplot(data = cancer_df, aes(x = pgr)) +
  geom_histogram(binwidth= 420, color= "#ce90b4", fill = "#b4006a") +
  theme(plot.title = element_text(size = 16), axis.title = element_text(size = 16), axis.text = element_text(size = 16)) +
  labs(x = "pgr",
       y = "Count",
       title = "pgr distribution")

ggarrange(p1_age, p2_nodes, p3_pgr, ncol =3)
```
<br>

```{r, fig.align='center', fig.width=15, fig.height=5}

p4_er <- ggplot(data = cancer_df, aes(x = er)) +
  geom_histogram(binwidth= 300, color= "#ce90b4", fill = "#b4006a") +
  theme(plot.title = element_text(size = 16), axis.title = element_text(size = 16), axis.text = element_text(size = 16)) +
  labs(x = "er",
       y = "Count",
       title = "er distribution")

p5_size <- ggplot(data = cancer_df, aes(x = size)) +
  geom_bar(width = 0.7, color= "#ce90b4", fill = "#b4006a") +
  geom_text(data = . %>% 
              group_by(size) %>% 
              tally() %>% 
              mutate(prop = round(n/sum(n),3)) %>% 
              ungroup(),
            aes(y = n, label = scales::percent(prop)),
            position = position_stack(vjust=0.5),
            colour = "white", size = 5)+
  theme(plot.title = element_text(size = 16), axis.title = element_text(size = 16), axis.text = element_text(size = 16)) +
  labs(x = "Size",
       y = "Count",
       title = "Size distribution")

p6_grade <- ggplot(data = cancer_df, aes(x = factor(grade))) +
  geom_bar(width = 0.7, color= "#ce90b4", fill = "#b4006a") +
  geom_text(data = . %>% 
              group_by(grade) %>% 
              tally() %>% 
              mutate(prop = round(n/sum(n),3)) %>% 
              ungroup(),
            aes(y = n, label = scales::percent(prop)),
            position = position_stack(vjust=0.5),
            colour = "white", size = 5)+
  theme(plot.title = element_text(size = 16), axis.title = element_text(size = 16), axis.text = element_text(size = 16)) +
  labs(x = "Grade",
       y = "Count",
       title = "Grade distribution")

ggarrange(p4_er, p5_size, p6_grade, ncol =3)

```
<br>

```{r, fig.align='center', fig.width=15, fig.height=5}

p7_meno <- ggplot(data = cancer_df, aes(x = factor(meno), color = factor(meno), fill = factor(meno))) +
  geom_bar(width = 0.7) +
  scale_x_discrete(labels = c("No", "Yes")) +
  scale_fill_manual(values = c("#b4006a", "peru")) +
  scale_color_manual(values = c("#ce90b4", "goldenrod2")) +
  geom_text(data = . %>% 
              group_by(meno) %>% 
              tally() %>% 
              mutate(prop = round(n/sum(n),3)) %>% 
              ungroup(),
            aes(y = n, label = scales::percent(prop)),
            position = position_stack(vjust=0.5),
            colour = "white", size = 5)+
  theme(plot.title = element_text(size = 16), axis.title = element_text(size = 16), axis.text = element_text(size = 16)) +
  theme(legend.position = "none") +
  labs(x = "Meno",
       y = "Count",
       title = "Meno distribution")


p8_hormon <- ggplot(data = cancer_df, aes(x = factor(hormon), color = factor(hormon), fill = factor(hormon))) +
  geom_bar(width = 0.7) +
  scale_x_discrete(labels = c("No", "Yes")) +
  scale_fill_manual(values = c("#b4006a", "peru")) +
  scale_color_manual(values = c("#ce90b4", "goldenrod2")) +
  geom_text(data = . %>% 
              group_by(hormon) %>% 
              tally() %>% 
              mutate(prop = round(n/sum(n),3)) %>% 
              ungroup(),
            aes(y = n, label = scales::percent(prop)),
            position = position_stack(vjust=0.5),
            colour = "white", size = 5)+
  theme(plot.title = element_text(size = 16), axis.title = element_text(size = 16), axis.text = element_text(size = 16)) +
  theme(legend.position = "none") +
  labs(x = "Hormon",
       y = "Count",
       title = "Hormon distribution")

p9_chemo <- ggplot(data = cancer_df, aes(x = factor(chemo), color = factor(chemo), fill = factor(chemo))) +
  geom_bar(width = 0.7) +
  scale_x_discrete(labels = c("No", "Yes")) +
  scale_fill_manual(values = c("#b4006a", "peru")) +
  scale_color_manual(values = c("#d09ea7", "goldenrod2")) +
  geom_text(data = . %>% 
              group_by(chemo) %>% 
              tally() %>% 
              mutate(prop = round(n/sum(n),3)) %>% 
              ungroup(),
            aes(y = n, label = scales::percent(prop)),
            position = position_stack(vjust=0.5),
            colour = "white", size = 5)+
  theme(plot.title = element_text(size = 16), axis.title = element_text(size = 16), axis.text = element_text(size = 16)) +
  theme(legend.position = "none") +
  labs(x = "Chemo",
       y = "Count",
       title = "Chemo distribution")

ggarrange(p7_meno, p8_hormon, p9_chemo, ncol = 3)

```
<br>

```{r, fig.align='center', fig.width=15, fig.height=5}

p10_recur <- ggplot(data = cancer_df, aes(x = factor(recur), color = factor(recur), fill = factor(recur))) +
  geom_bar(width = 0.7) +
  scale_x_discrete(labels = c("No", "Yes")) +
  scale_fill_manual(values = c("#b4006a", "peru")) +
  scale_color_manual(values = c("#ce90b4", "goldenrod2")) +
  geom_text(data = . %>% 
              group_by(recur) %>% 
              tally() %>% 
              mutate(prop = round(n/sum(n),3)) %>% 
              ungroup(),
            aes(y = n, label = scales::percent(prop)),
            position = position_stack(vjust=0.5),
            colour = "white", size = 5)+
  theme(plot.title = element_text(size = 16), axis.title = element_text(size = 16), axis.text = element_text(size = 16)) +
  theme(legend.position = "none") +
  labs(x = "Recur",
       y = "Count",
       title = "Recur distribution")

p11_death <- ggplot(data = cancer_df, aes(x = factor(death), color = factor(death), fill = factor(death))) +
  geom_bar(width = 0.7) +
  scale_x_discrete(labels = c("No", "Yes")) +
  scale_fill_manual(values = c("#b4006a", "peru")) +
  scale_color_manual(values = c("#ce90b4", "goldenrod2")) +
  geom_text(data = . %>% 
              group_by(death) %>% 
              tally() %>% 
              mutate(prop = round(n/sum(n),3)) %>% 
              ungroup(),
            aes(y = n, label = scales::percent(prop)),
            position = position_stack(vjust=0.5),
            colour = "white", size = 5)+
  theme(plot.title = element_text(size = 16), axis.title = element_text(size = 16), axis.text = element_text(size = 16)) +
  theme(legend.position = "none") +
  labs(x = "Death",
       y = "Count",
       title = "Death distribution")

ggarrange(p10_recur, p11_death,  ncol = 3)

```
<br>

**Survival data representations and censoring**

In survival analysis, the interest is on analyzing data where the outcome variable is time to an event, measured from a defined starting point. In this analysis, the outcome variable is `dtime` which is the duration of survival or the days to death or last follow-up of breast cancer patients. The event of interest is death as indicated by the variable `death` (class 1/yes - dead) and the starting point is the patient's surgery indicated by the year of surgery (`year`)and age of patient at surgery (`age`).  

The figure below shows a histogram of `dtime` and count of patients based on `death` status for the study period.

```{r, fig.align="center", fig.width=10, fig.height=5}
hist_01 <- ggplot(cancer_df, aes(dtime)) +
  geom_histogram(aes(fill = deathF), alpha = 0.8, bins = 40, position="identity") +
  scale_fill_manual(values = c("#b4006a", "peru"), labels = c("0 - Alive/Censored", "1 - Dead")) +
  labs(title = " Distribution of survival status (death)") +
  theme(plot.title = element_text(size=12))

hist_01
```
<br>

In the dataset, censoring occurs when information about a patient's survival is incomplete based on the available data. Since there is no missing value in the `year`, it can be said that there can be right censoring in the dataset, i.e., some patients may not have died  by the end of study period (alive), some patients might have been lost to follow up or have died after the study period (censored). These can be seen in the graphical representations of survival data below.

The figure below shows survival data of randomly selected 30 breast cancer patients from the dataset. The survival data is represented using the triplet - entry age, exit age, event. From the starting point which was the `age` at surgery, the patient was followed until they experience the event (death) or they are censored. If the patient had a recurrence (1-relapse) within the study period, it is marked marked by a 'purple' dot.   

```{r, fig.align="center", fig.width= 12, fig.height=8}
set.seed(300)
cancer_df_sample <- cancer_df %>% 
  slice_sample(n=30) %>% 
  mutate(order = pid,
         start = 0,
         age2 = ((age*365) + dtime)/365,
         age_relapse = ((age*365)+ rtime)/365)

ggplot(data=cancer_df_sample) +
  geom_point(aes(x = age, y=factor(order)), color = "black", size = 2) +
  geom_segment(aes(x=age, xend = age2, y=factor(order), yend = factor(order)), size =1.2) +
  geom_point(aes(x=age_relapse, y=factor(order)), color = "purple", size = 2) +
  geom_point(aes(x=age2, y=factor(order), shape = deathF, color = deathF), size = 3, stroke = 1.5) +
  scale_x_continuous("Age (years)", breaks = seq(20, 90, 10), limits = c(20, 90)) +
  scale_y_discrete("Patient ID") +
  scale_shape_manual(NULL, values = c(20,4), labels = c("Alive/Censored", "Died")) +
  scale_color_manual(NULL, values = c("peru", "red"), labels = c("Alive/Censored", "Died")) +
  theme_gray()

```

<br>

Another way to present survival data of the 30 sampled patients is using the pair: follow-up (survival) time, event. This is shown in the figure below.

```{r, fig.align="center", fig.width= 12, fig.height=8}

ggplot(data=cancer_df_sample) +
  geom_segment(aes(x=start, xend = dtime, y=factor(order), yend = factor(order)), size =1.2) +
  geom_point(aes(x=rtime, y=factor(order)), color = "purple", size = 2) +
  geom_point(aes(x=dtime, y=factor(order), shape = deathF, color = deathF), size = 3, stroke = 1.5) +
  geom_text(aes(x=dtime, y=factor(order),label = dtime), nudge_x = 200, size = 4) +
  scale_x_continuous("Survival time (days)") +
  scale_y_discrete("Patient ID") +
  scale_shape_manual(NULL, values = c(20,4), labels = c("Alive/Censored", "Died")) +
  scale_color_manual(NULL, values = c("peru", "red"), labels = c("Alive/Censored", "Died")) +
  theme_gray()

```

```{r, results='hide'}
# get the longest time for alive/censored (death = 0) (recur = 0)
max.time.0 <-cancer_df %>% filter(death == 0) %>% filter(recur == 0) %>% filter(dtime == max(dtime))
a0 <- max.time.0 %>% mutate(numyears = dtime/365) %>% mutate(recuryears = rtime/365)%>% select(dtime, numyears, recuryears, year, age, recur)

# get the longest time for dead (death = 1) (recur = 0)
max.time.1 <-cancer_df %>% filter(death == 1) %>% filter(recur == 0) %>% filter(dtime == max(dtime))
a1 <- max.time.1 %>% mutate(numyears = dtime/365) %>% mutate(recuryears = rtime/365)%>% select(dtime, numyears, recuryears, year, age, recur)

rbind(a1, a0)

```
Based on the dataset:

* Without recurrence: The longest time until death (death = 1) was 5830 days (about 16 years) after surgery by a patient who had surgery in 1983 at 66 years old (recur = 0)
* Without recurrence: The longest time recorded for alive/censored (death = 0) was 7043 days (about 19 years) after surgery by a patient who had surgery in 1981 at 49 years old (recur = 0).

```{r, results='hide'}
# get the longest time for alive/censored (death = 0)(recur = 1)
max.time.0 <-cancer_df %>% filter(death == 0) %>% filter(recur == 1) %>% filter(dtime == max(dtime))
a0 <- max.time.0 %>% mutate(numyears = dtime/365) %>% mutate(recuryears = rtime/365) %>% select(dtime, rtime, numyears, recuryears, year, age, recur)

# get the longest time for dead (death = 1)(recur = 1)
max.time.1 <-cancer_df %>% filter(death == 1) %>% filter(recur == 1) %>% filter(dtime == max(dtime))
a1 <- max.time.1 %>% mutate(numyears = dtime/365) %>% mutate(recuryears = rtime/365) %>% select(dtime, rtime, numyears, recuryears, year, age, recur)

rbind(a1, a0)

```

* With recurrence: The longest time until death (death = 1) was 6233 days (about 17 years) after surgery by a patient who had surgery in 1983 at 64 years old. The patient had recurrence 3732 days (about 10 years) after surgery (recur = 1)
* With recurrence: The longest time recorded for alive/censored (death = 0) was 6886 days (about 19 years) after surgery by a patient who had surgery in 1980 at 45 years old. The patient had recurrence 2570 days (about 7 years) after surgery (recur = 1)


```{r, results='hide'}
# get the shortes time for alive/censored (death = 0)(recur = 0)
min.time.0 <-cancer_df %>% filter(death == 0) %>% filter(recur == 0) %>% filter(dtime == min(dtime))
a0 <- min.time.0 %>% mutate(numyears = dtime/365) %>% mutate(recuryears = rtime/365)%>% select(dtime, numyears, recuryears, year, age, recur)

# get the shortest time for dead (death = 1)(recur = 0)
min.time.1 <-cancer_df %>% filter(death == 1) %>% filter(recur == 0) %>% filter(dtime == min(dtime))
a1 <- min.time.1 %>% mutate(numyears = dtime/365) %>% mutate(recuryears = rtime/365)%>% select(dtime, numyears, recuryears, year, age, recur)

rbind(a1, a0)

```

Unfortunately,

* Without recurrence: The shortest time until death (death = 1) was 45 days after surgery by a patient who had surgery in 1987 at 75 years old (recur = 0)
* Without recurrence: The longest time recorded for alive/censored (death = 0) was 36 days after surgery by a patient who had surgery in 1992 at 36 years old (recur = 0). The patient was probably censored, loss to follow-up.

```{r, results='hide'}
# get the shortest time for alive/censored (death = 0)(recur = 1)
min.time.0 <-cancer_df %>% filter(death == 0) %>% filter(recur == 1) %>% filter(dtime == min(dtime))
a0 <- min.time.0 %>% mutate(numyears = dtime/365) %>% mutate(recuryears = rtime/365) %>% select(dtime, rtime, numyears, recuryears, year, age, recur)

# get the shortest time for dead (death = 1)(recur = 1)
max.time.1 <-cancer_df %>% filter(death == 1) %>% filter(recur == 1) %>% filter(dtime == min(dtime))
a1 <- max.time.1 %>% mutate(numyears = dtime/365) %>% mutate(recuryears = rtime/365) %>% select(dtime, rtime, numyears, recuryears, year, age, recur)

rbind(a1, a0)

```

* With recurrence: The shortest time until death (death = 1) was 141 days after surgery by a patient who had surgery in 1988 at 68 years old. The patient had recurrence 76 days after surgery (recur = 1)
* With recurrence: The longest time recorded for alive/censored (death = 0) was 129 days after surgery by a patient who had surgery in 1990 at 53 years old. The patient had recurrence 87 days after surgery (recur = 1). The patient was probably censored, loss to follow-up.

<br>
<br>

### III. Kaplan-Meier Survival Curves and Log Rank Tests

**A. Kaplan-Meier Curves**

Kaplan- Meier curves are non-parametric summary of the probability of survival past a certain time. The KM survival curve of the breast cancer patients from the dataset is shown below:

```{r, echo=TRUE}
mod01_sfit <- survfit(Surv(dtime, death) ~ 1, data = cancer_df)
```

```{r, fig.align="center", fig.width= 9, fig.height= 4}
ggsurvplot(mod01_sfit, 
           conf.int = TRUE,
           risk.table = TRUE, fontsize = 4,
           risk.table.col = "strata",
           linetype = "strata",
           xlim = c(0, 7043),
           surv.median.line = "hv",
           ggtheme = theme_bw(), 
           palette = "#b4006a")
```
```{r}
surv_median(mod01_sfit)
```
The median survival time for the breast cancer patients in the dataset is 4033 days.

Using the KM curve, the 5-year survival rate for primary breast cancer based on the available data in this dataset is 74.4% and the 10-year survival rate is 55.2% as shown below.

```{r, echo = TRUE}
summary(mod01_sfit, times = 1825)    #1825 days in 5 years
summary(mod01_sfit, times = 3650)    #3650 days in 10 years
```
<br>

Separate KM curves can also be plotted and compared for different variables to explore if different values/levels of the variables affect survival probabilities. Below are the KM curves for the features `age.group`, `menoF`, `nodes.group`, `hormonF`, `size`, `chemoF`, `gradeF`, and `recurF`.

```{r, echo=FALSE, fig.align="center", fig.width= 14, fig.height= 18}
m1_age <- survfit(Surv(dtime, death) ~ age.group, data = cancer_df)
m1_nodes <- survfit(Surv(dtime, death) ~ nodes.group, data = cancer_df)
m1_pgr <- survfit(Surv(dtime, death) ~ pgr, data = cancer_df)
m1_er <- survfit(Surv(dtime, death) ~ er, data = cancer_df)

m1_size <- survfit(Surv(dtime, death) ~ size, data = cancer_df)
m1_grade <- survfit(Surv(dtime, death) ~ gradeF, data = cancer_df)
m1_meno <- survfit(Surv(dtime, death) ~ menoF, data = cancer_df)
m1_hormon <- survfit(Surv(dtime, death) ~ hormonF, data = cancer_df)
m1_chemo <- survfit(Surv(dtime, death) ~ chemoF, data = cancer_df)
m1_recur <- survfit(Surv(dtime, death) ~ recurF, data = cancer_df)

# define titles
titlelist <- list("Age", "Nodes", "pgr", "er", "Size", "Grade", "Meno", "Hormon", "Chemo", "Recur")

#Plot

survplots <- ggsurvplot(list(m1_age, m1_nodes, m1_size, m1_grade, m1_meno, m1_hormon, m1_chemo, m1_recur),
           pval = TRUE, pval.method = TRUE, conf.int = TRUE,
           risk.table = TRUE, fontsize = 5,
           risk.table.col = "strata",
           linetype = "strata",
           xlim = c(0, 7043),
           surv.median.line = "hv",
           ggtheme = theme_bw(), 
           palette = c("#b4006a", "peru", "green4"),
           risk.table.height = 0.30)

arrange_ggsurvplots(survplots, print = TRUE, ncol = 2, nrow = 4)

```
<br>

**B. Log Rank test**

Log rank test was used to determine if differences in the survival curves of each feature/variable were significant. For instance, using the log rank test for `gradeF`:

```{r, echo=TRUE}
log.rank.grade <- coxph(Surv(dtime, death) ~ gradeF, data = cancer_df)
summary(log.rank.grade)
```

The result shows that the survival curves for `gradeF` are significantly different for at least some time intervals (low p-value). The median survival time for those who have tumor differentiation grade 2 was 5653 days while 3700 days for those with tumor grade 3 (shown below and also in the KM curve).

```{r}
surv_median(m1_grade )
```

```{r, results='hide'}
# create a comparison table

covariates <- c("age.group", "nodes.group", "pgr", "er", "size", "gradeF", "menoF", "hormonF", "chemoF", "recurF")

surv.formulas <- sapply(covariates, 
                        function(x) as.formula(paste('Surv(dtime, death) ~ ', x)))

log.rank.fits <- lapply(surv.formulas,
                        function(x) {coxph(x, data= cancer_df)})

# extract log-rank test results
log.rank.results <- lapply(log.rank.fits,
                           function(x) {
                             x <- summary(x)
                             results <- c(round(x$sctest[1], digits = 4), x$sctest[2], round(x$sctest[3], digits = 6))
                             names(results) <- c("Score (logrank) test", "df", "p.value")
                             return(results)
                           })

results.trans <- t(as.data.frame(log.rank.results))
results.df <- results.trans %>% 
  as.data.frame() %>% 
  mutate(Remark = ifelse(p.value < 0.05, "The survival curves have statistically significant differences.", 
                         "The survival curves have statistically non-significant differences")) %>% 
  print()
```
<br>

The table below shows a summary of the log rank test results performed on the features or covariates. The covariates `er` (estrogen receptor biomarker) and `chemoF` (indicator of whether the patient received chemotherapy) both have KM survival curves that have statistically non-significant firrences. 

```{r}
sjPlot::tab_df(results.df, show.rownames = TRUE, digits = 4)

```
<br>
<br>

### IV. Cox proportional hazard regression models

Cox regression is used for modelling time-to-event data to relate the outcome (survival duration or `dtime`) to one or more covariates.   

For this anaysis, three Cox regression models were evaluated

1. Model 1 - Reduced multivariable model
2. Model 2 - Full multivariable model
3. Model 3 - Stratified model

Note that Model 3 was  developed as a result of the diagnostics test after comparing Models 1 and 2. The models are detailed below.

<br>

**Model 1 - Reduced multivariable model**

This model fits a Cox regression model  that relates time-to-event (`dtime`, `death`) to a reduced list of relevant covariates, i.e., excluding `er` and `chemoF` in the model. The basis of this was the results of the log rank tests performed above.

```{r, echo=TRUE}
surv.model <- coxph(Surv(dtime, death) ~ age.group + nodes.group + pgr + size +
                      gradeF + menoF + hormonF + recurF, data = cancer_df)
summary(surv.model)
```

<br>

**Model 2 - Full multivariable model**

This is a more complex model that fits a Cox regression model which relates time-to-event (`dtime`, `death`) to all relevant covariates from the dataset.

```{r, echo=TRUE}
surv.model_full <- coxph(Surv(dtime, death) ~ age.group + nodes.group + pgr + er + size +
                      gradeF + menoF + hormonF + chemoF + recurF, data = cancer_df)

summary(surv.model_full)
```

<br>

**Comparing Models 1 and 2**

A likelihood ratio test was used to compare nested models 1 and 2 (below). The result showed preference for the reduced and simpler model (Model 1) providing adequate fit (based on p-value).

```{r}
# anova test
anova(surv.model, surv.model_full)
```
<br>

**Model Diagnostics**

A key assumption for the Cox regression model is proportional hazards. A statistical test based on Schoenfeld residuals was used to check if Model 1 violates the proportionality assumption..

```{r, echo=TRUE}
stat.test <- cox.zph(surv.model)
stat.test
```
The result above showed that `age.group` and `pgr` variables violated the proportional hazard assumption (based on individual and global p-values <0.05). Plots of Schoenfeld residuals against `dtime` aare shown below:

```{r, fig.align='center', fig.width=15, fig.height= 10}
ggcoxzph(stat.test)

```

<br>

**Model 3 - Stratified model**

This model fits a Cox regression model that attempts to resolve the violation of proportionality assumption of Model 1 by stratification, i.e., splitting the data of violating covariates (`age.group` and `pgr`) into groups termed strata.  

```{r}
surv.model.stratified <- coxph(Surv(dtime, death) ~ strata(age.group) + nodes.group + strata(pgr) + size +
                      gradeF + menoF + hormonF + recurF, data = cancer_df)

summary(surv.model.stratified)
```

<br>

Result of the statistical test as shown below (all p-values >0.05) confirmed that the Model 3 resolves the violation and selected as the  best-fitting model.  

```{r, echo=TRUE}
cox.zph(surv.model.stratified)
```
<br>

**Hazard Ratio**

The Cox regression model gives estimates of the  **hazard ratios**. The hazard rate describes, in this analysis, the relative probability of `death` occuring at `dtime`, given that `death` hasn't happened up until time `dtime`. 

The figure below shows the forest plot for Model 3 presenting the hazard ratios of the covariates for breast cancer patients in the dataset. 

```{r, fig.align='center', fig.width=16, fig.height=14}
attr(surv.model.stratified$terms, "dataClasses") <- attr(surv.model.stratified$terms, "dataClasses")[-c(1, 4)]
ggforest(model = surv.model.stratified, data = cancer_df, fontsize = 1.7)
```
<br>

The following can be interpreted:

* In general, every hazard ratio (HR) represents a relative risk of death that compares one of the categories of the covariate to the baseline “reference”. A HR <1 means that there is a decreased hazard (or risk) of death (the outcome event) and a HR >1 means that there is an increased hazard (or risk) of death, relative to the reference. Hence, this also implies that if the 95% confidence interval for HR include the value ‘1’, it means that the resulting p-value will be high (> 0.05) which would mean that the difference to the baseline “reference” is not significant. This is true for `nodes.group` 21-34 category, `gradeF` grade 3 category, and `hormonF` Yes category.

* In terms of number of positive lymph nodes (`nodes.group`): After adjusting to other variables, the hazard of dying is 79% higher for patients with 11-20 positive nodes compared to those with 0-10 positive nodes (HR = 1.79, 95% CI: 1.4-2.3).

* In terms of tumor size range (`size`): After adjusting to other variables, the hazard of dying is 52% higher for patients with tumor size range of 20-50 compared to those with tumor size range of <= 20 (HR = 1.52, 95% CI: 1.29-1.8). The hazard of dying is 124% higher for patients with tumor size range of >= 50 compared to those with tumor size range of <= 20 (HR = 2.24, 95% CI: 1.78-2.8).

* In terms of patient's menopausal status (`menoF`): After adjusting to other variables, the hazard of dying is 23% higher for patients postmenopausal (1-Yes) compared to those premenopausal (0-No) (HR = 1.23, 95% CI: 1.01-1.5).

* In terms of whether the patient had a recurrence or not (`recurF`): After adjusting to other variables, the hazard of dying is very high, about 888% higher (almost 9x higher) for who had relapse compared to those who did not (HR = 9.88, 95% CI: 7.94-12.3).

<br>
<br>

### V. Conclusion

In this project, the survival rate of primary breast cancer patients were determined based on the available data in the dataset. The hazard ratios for each variable (covariate) were also estimated with a stratified Cox reggression model. From the results, the number of positive lymph nodes, tumor size, menaupausal status and recurrence (relapse) are the main factors that have significant effect on the survival rate of the patients.  

<br>
<br>
<br>
<br>




